\vspace{1.0cm} \LARGE \bf Genomic horizon

### 2019-03-18 genomic-horizon.tex
Text changes, finishing up.

@@

\author[1$\ast$]{Gytis Dudas}
\author[1]{Trevor Bedford}

\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA}

\begin{document}

@@

\begin{abstract}
Inexpensive pathogen genome sequencing has transformed the field of genetic epidemiology into genomic epidemiology.
Outbreak investigations have benefited the most from the inclusion of genome sites beyond short and rapidly evolving genes.
Despite the reduction in cost, pathogen genome sequencing has not seen universal adoption across pathogen-specific fields.
In this study we use a well-characterised Ebola virus dataset to show the differences in phylogenetic, temporal, and spatial resolution between complete genomes ($\sim$19kb long) and the rapidly evolving glycoprotein (GP) gene ($\sim$2kb long).
By quantifying errors in phylogenetic reconstruction we hope to persuade researchers looking to sequence pathogens during outbreaks of the utility of complete genome sequencing.
We further generalise our results by relating four parameters - probability, evolutionary rate, alignment length and waiting times - that determine how much information can be encoded by evolving pathogen lineages via mutations, which can help guide future sequencing studies.
\end{abstract}

@@

\section*{Introduction}
The combination of decreasing cost of sequencing and the unparalleled insight it offers have led to the adoption of pathogen genetic sequencing as one of the most effective tools in the molecular epidemiologist's toolkit.
When coupled with sophisticated models of evolution, pathogen sequences can be used to look into epidemiological features such as cryptic transmission \citep{faria_establishment_2017}, migration \citep{dudas_virus_2017}, and origins \citep{smith_origins_2009} of infectious diseases, amongst others.
Pathogen sequences also contain information about past temporal dynamics beyond the time span over which sequences have been collected \citep{raghwani_origin_2012} owing to inheritance of mutations by future generations.

Molecular phylogenetic approaches rely on decoding these patterns of shared mutations into a nested data structure known as the phylogenetic tree.
Population structure is often discernible in phylogenies, where specific lineages either arise via mutation or arrive from elsewhere and come to dominate within a particular demographic compartment (tissues, hosts, geographic areas, \textit{etc}.).
Pathogens often have short generation times and low replication fidelity (RNA viruses) such that mutations are generated at a rapid pace \citep{drummond_measurably_2003,biek_measurably_2015}, thus differentiating lineages genetically as they spread.
With appropriate sampling and information (``metadata'') about sequences, historical shifts in population structure can be inferred and quantified from pathogen phylogenies.

Before widespread adoption of high-throughput sequencing limitations and costs led to amplification and sequencing of short fragments of pathogen genomes \citep{jin_genetic_1999,jin_proposal_2005}.
These subgenomic fragments were often chosen for their diversity, like viral surface glycoproteins evolving neutrally or under diversifying selection, or their utility, such as resistance profiling for HIV \textit{pol} \citep{kaye_phylogenetic_2008,rhee_human_2003}.
Whilst subgenomic fragments of pathogens are very accurate and specific as diagnostic markers, their length (dictated by the compromise between information content and ease of sequencing) limits their utility in detailed molecular epidemiology investigations \citep{wohl_co-circulating_2018}, as only mutations occurring with the small region of the genome are available for phylogenetic inference.

Mutations, prior to the action of selection, occur randomly across a genome via misincorporation of nucleotides during replication by the polymerase.
Many of these mutations are filtered out by selection if they adversely affect protein function, with leftover mutations being beneficial, neutral, or mildly deleterious.
Neutral pathogen variation at the nucleotide level ebbs and flows with forces of population genetics but outnumbers beneficial variation that tends to sweep rapidly.
Mutations are often modelled as a Poisson process \citep{jukes_evolution_1969}, and the probability of observing a mutation at a single site is exponentially distributed.
The probability of 0 mutations at a single site is related to the waiting time \textit{t} and the rate at which mutations are observed in sequences \textit{R} (evolutionary rate): $e^{-Rt}$.
The probability of at least one mutation at a single site after time \textit{t} under an evolutionary rate \textit{R} is therefore:

\begin{equation}
  P = 1-e^{-Rt}
\end{equation}

Higher evolutionary rates \textit{R} or waiting times \textit{t} result in higher probabilities of observing at least one mutation. Since sites are treated as evolving independently, the probability of observing a mutation across more than one site is related to the number of sites \textit{L} under observation:

\begin{equation}
  P = 1-e^{-RLt}
\end{equation}

Where -\textit{RL} corresponds to the raw rate of sampled mutations expressed as substitutions per year. Under a given evolutionary rate \textit{R}, probability of a mutation p, and sequence length \textit{L}, we can rearrange the equation to quantify the waiting times for mutations:

\begin{equation}
  t = \frac{log(P)}{-RL}
\end{equation}

When the evolutionary rate \textit{R}, sequence length \textit{L}, or probability of observing a mutation p are low waiting times are lengthened and vice versa.
Since both maximum plausible evolutionary rates and genome length are dictated by deleterious mutation load \citep{gago_extremely_2009}, neither quantity will vary substantially for a given pathogen, especially viruses.
Because of these natural constraints there is thus a lower limit to how many mutations can be expected to occur in a given length of sequence evolving at a particular evolutionary rate.
The mean waiting time for a mutation is easily calculated as 1/\textit{RL}.
We refer to this mean waiting time as the temporal horizon, which reflects a theoretical temporal resolution over which mutations arise and are available for phylogenetic inference in a system.

We use this study to argue in favour of sequencing complete pathogen genomes in outbreak scenarios where lineages need to be tracked at a fine scale.
We base our argument on the fact that pathogen genomes have an inherent mutation/temporal resolution limit which is already substantial for fine scale epidemiology and that sequencing subgenomic fragments of a pathogen results in loss of information that can make sequence data unreliable beyond diagnostics and typing.
We rely on a subset of a well characterised \citep{dudas_virus_2017} dataset of Ebola virus genomes from West Africa collected in 2014-2015.
By analysing both complete viral genomes and the same dataset reduced to just the surface glycoprotein GP (corresponding to roughly 10\% of all available sites), we show substantial loss of predictive power, incorrect or unreliable reconstructions of past events, and migration model, despite increased evolutionary rate in GP compared to complete genome.
We hope that analyses presented in this study serve as a guide for researchers doing genetic epidemiology and will be of interest to model developers too.

\begin{figure}[h]
 \centering
	\includegraphics[width=0.75\textwidth]{figures/fig1_trees.png}
	\caption{\textbf{Phylogenies of West African Ebola virus genomes (top) or GP sequences (bottom).}
	Temporal phylogenies recovered using BEAST are shown on the left, maximum likelihood phylogenies recovered with RAxML are on the right.
  Tips are coloured based on country (Sierra Leone in blue, Liberia in red, Guinea in green) and location (lighter colours indicate administrative divisions lying towards west of the country).
  Tips outlined in white indicate the 60 chosen for date and location masking.
  In temporal phylogenies branches are also coloured based on GLM-inferred ancestral locations.
  Nodes in temporal phylogenies with <0.10 posterior probability are indicated with grey X marks.
  Maximum likelihood phylogenies on the right are rooted via temporal regression in TreeTime.
	}
	\label{trees}
\end{figure}

@@

\subsection*{Loss of phylogenetic signal}
In the past amplifying smaller fragments of pathogen genomes was often a pragmatic compromise between the cost of sequencing and information content, leading to sequencing of short but rapidly evolving pathogen genes or genomic regions.
What few sequences were available were often collected months apart or across large geographic areas, ensuring that many sequences were genetically distinct either via mutations occurring in the intervening time or through prior population differentiation.
However, in outbreak scenarios shorter genome fragments often do not accumulate mutations at a pace rapid enough to be of much use over short timescales.

Fig \ref{trees} shows the reconstructed phylogenies in substitution space (right) and time space (left) for 600 complete Ebola virus genomes (top) or just GP sequences (bottom).
Although higher levels of divergence are observed in the GP dataset, as seen from tree height, the differences in the number of non-polytomic nodes between genomic and GP data are clear, indicating substantially better resolution in disentangling the exact relationships between lineages in the former.
Internal branches of a phylogeny correspond to hypotheses of common ancestry and in the case of GP only 42 internal nodes are identified in the maximum likelihood phylogeny compared to 210 internal nodes for complete genomes.
Similarly, broad aspects of the West African epidemic can be inferred from both GP and genome phylogenies, such as Ebola virus' origins in Guinea, but the details of how it spread onwards are largely lost in the GP phylogeny.

Unlike maximum likelihood phylogenies where branch lengths are directly proportional to the expected number of substitutions, branch lengths in temporal phylogenies are usually smoothed out by the fact that a range of dates are compatible with a given number of mutations on a branch.
Thus even large polytomies can be resolved into a branching structure derived from the tree prior, albeit without much support for any given configuration.
This can be seen in Fig \ref{trees}, where temporal phylogenies (left) appear to have similar degrees of resolution, yet the GP dataset (bottom) contains more nodes with less than 0.10 posterior support (marked by grey crosses).
Similarly, there is a noticeable degree of branch clustering by country in the GP temporal phylogeny, possibly caused by proximity of locations within country, which in the absence of genetic information cannot be resolved to the same degree as with genomic data.

\begin{figure}[h]
 \centering
	\includegraphics[width=0.75\textwidth]{figures/fig2_dates.png}
	\caption{\textbf{Masked tip date inference from genomes (left) and GP sequences (right).}
  Inferred collection dates in the masked set based on genomes (red, left) and GP sequences (blue, right).
  Each vertical line corresponds to the 95\% highest posterior density for the inferred tip date (y-axis), coloured red (genome) or blue (GP) if it falls within the true collection date (x-axis) and black otherwise.
  Dashed diagonal line indicates the 1-to-1 line.
  Posterior distributions of residuals from the 1-to-1 line are shown at the bottom with the mean of the distribution indicated by a tick mark.
	}
	\label{dates}
\end{figure}

@@

Inferring masked tip dates from 10\% of the sequences (Fig \ref{dates}) is possibly one of the most intuitive ways to show both the noisiness of molecular clock estimates, as reflected in the width of 95\% highest posterior density intervals for inferred dates, but also the differences between GP and genome alignments.
True collection dates for genomes are mostly (56 out of 60) within the 95\% highest posterior density of estimated dates, and the absolute posterior error is around 0.081 years, corresponding to roughly a month of error across all masked tips.
The 95\% HPDs for inferred dates in the GP dataset, on the other hand, manages to capture more of the true dates (58 out of 60) at the cost of markedly reduced precision, with absolute posterior error going up to 0.417 years (or close to 5 months).
Absolute error for both genome and GP datasets is relatively close to theoretical expectations: one month versus predicted 3 weeks for genomes, and 5 months versus expected 3-6 months.
For many tips in the GP dataset independent Markov chains converged onto different dates for masked tips (\textit{i.e.} local maxima) resulting in multi-peaked posterior samples after combining independent analyses.

\begin{figure}[h]
 \centering
	\includegraphics[width=0.75\textwidth]{figures/fig3_locations.png}
	\caption{\textbf{Masked tip location inference from genomes (left) and GP sequences (right).}
  Horizontal bars indicate the posterior distribution of masked tip locations, coloured by country (Sierra Leone in blue, Liberia in red, Guinea in green) and location (lighter colours indicate administrative divisions lying towards west of the country).
  The correct location of each tip is outlined in white, with the smaller plot to the right indicating the posterior probability of the correct location.
  Posterior distribution of probability-weighted distances between population centroids of inferred and correct locations is shown at the bottom with mean indicated by a tick mark.
	}
	\label{locations}
\end{figure}

\begin{figure}[h]
 \centering
  \subfloat{
  \includegraphics[width=0.45\textwidth]{figures/figXA_histories.png}
  }
  \subfloat{
  \includegraphics[width=0.45\textwidth]{figures/figXB_histories.png}
  }
  \\
  \subfloat{
  \includegraphics[width=0.45\textwidth]{figures/figXC_histories.png}
  }
  \subfloat{
  \includegraphics[width=0.45\textwidth]{figures/figXD_histories.png}
  }
  \caption{\textbf{Posterior traces of ancestral locations and posterior migrations for four Ebola virus lineages from genomes (top) and GP sequences (bottom).}
  The inferred ancestral branch location is logged at time points along the path from selected tips to the root of the tree, across the posterior distribution of trees.
  The smoothed trajectories are an indication of where and when a lineage that gave rise to a particular tip existed.
  Maps on the right show migration events that are inferred to have taken place, coloured by their posterior probability, with migrations with <0.05 posterior support are shown as dotted lines.
  All lineages trace back to Gu\'{e}ck\'{e}dou prefecture of Guinea, where the original zoonotic transmission event occurred near the Guinean border with Sierra Leone and Liberia.
  Some lineages are also descended from an early spillover event into Sierra Leone.
  }
	\label{trace}
\end{figure}

\subsubsection*{Loss of power for phylogeographic reconstruction}

Differences between genomic and GP datasets are clear and dramatic when looking at both phylogenies (Fig \ref{trees}) and masked date inference (Fig \ref{dates}), but less pronounced when trying to infer the location of a masked tip (Fig \ref{locations}).
Although locations are correctly inferred more often and with greater support in genomic sequences compared to just the GP gene, there are numerous tips whose locations are not correctly inferred from genome sequences.

This might reflect the nature of these parameters of interest, since phylogenies and date inference ultimately draw information from mutation accumulation via relatively straightforward models of sequence evolution with limited parameter space.
In contrast, migration is a far more complicated and nuanced process without a de facto standard for modelling, though continuous time Markov chain (CTMC) approaches are widely used.
Despite lack of an overwhelmingly obvious contrast between inference of masked location between genomes and GP sequences, cross entropies indicate much better performance with the former, at 6816.772 and 11831.626, respectively.
Similarly, locations are inferred correctly more often with complete genomes than with GP sequences (33 and 18, respectively).
Probability-weighted distances to the correct population centroid are also reduced for genomic data, at a posterior mean of 2.870 kilometres compared to 4.315 km for GP sequences.

In addition to assessing how well tip locations can be inferred from genetic information we also looked at how well historical patterns were reconstructed from sequence data.
To accomplish this we looked at the posterior distribution of ancestral locations for four tips with distinct histories.
These are EBOV|14859\_EMLK|KU296438|SLE|Kambia|2015-04-30, a lineage that was part of the initial sweep of Ebola virus through Sierra Leone via Kailahun and Kenema districts which ended up in Conakry in Guinea from where it jumped back into Sierra Leone late in the epidemic.
EBOV|EM\_004422|KR817097|GIN|Kissidougou|2015-01-02, which migrated through Kailahun district of Sierra Leone too, but then spread into Liberia from which it spilled back into Guinea via Macenta until being sequenced from a patient in Kissidougou, Guinea.
Similar to other lineages, EBOV|MK3462|KU296565|SLE|Bombali|2015-03-26 was part of the sweep through Sierra Leone, though unlike others remained in-country, migrating to the environs of Freetown and then Bombali district where it was sequenced.
EBOV|PL5294|KU296764|SLE|Kambia|2015-03-24, unlike others, belonged to a lineage that was not part of the Sierra Leonean sweep and instead part of an unusual and under-sampled lineage endemic to western Guinea, from where it spilled into Sierra Leone's Kambia district late in the epidemic.

>>>>

%%% Comments from lab meeting:
% - HIV in contour plot
% - time variation across a single mutation in RTT
% - review distances*probability
% - location inference - is the true location in the 95% set?
% - bias in collection impact on correlates of migration
% - mention that GLM is broad scale patterns and easy to get at
% - more focus on benefits of genomic epi
% - talk about impact of unsampled diversity i.e. whole clades missing
% - ancestry plots for all tips/summary across all tips. masked tip ancestry?
% - bar plot of mean waiting times between whole genomes and subgenomic fragments

@@

\author[1,2$\ast$]{Gytis Dudas}
\author[1]{Trevor Bedford}

\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA}
\affil[2]{Gothenburg Global Biodiversity Center, Gothenburg, Sweden}
\begin{document}

@@

\begin{abstract}
% Inexpensive pathogen genome sequencing has transformed the field of genetic epidemiology into genomic epidemiology.
Inexpensive pathogen genome sequencing has had a transformative effect on the field of phylodynamics, where ever increasing volumes of data have promised real-time insight into outbreaks of infectious disease.
As well as the sheer volume of pathogen isolates being sequenced, the sequencing of whole pathogen genomes, rather than select loci, has allowed phylogenetic analyses to be carried out at finer time scales, often approaching serial intervals for infections caused by rapidly evolving RNA viruses.
Despite its utility, whole genome sequencing of pathogens has not been adopted universally and targeted sequencing of loci is common in some pathogen-specific fields.
In this study we aim to highlight the utility of sequencing whole genomes of pathogens by re-analysing a well-characterised collection of Ebola virus sequences in the form of complete viral genomes ($\sim$19kb long) or the rapidly evolving glycoprotein (GP, $\sim$2kb long) gene.
We quantify changes in phylogenetic, temporal, and spatial inference resolution as a result of this reduction in data and compare these to theoretical expectations.
We propose a simple intuitive metric for quantifying temporal resolution, \textit{i.e.} the time scale over which sequence data might be informative of various processes as a quick back-of-the-envelope calculation of statistical power available to molecular clock analyses.
% We quantify the changes in phylogenetic, temporal, and spatial inference resolution available to researchers when complete virus genomes are used ($\sim$19kb long) versus just the rapidly evolving glycoprotein (GP) gene ($\sim$2kb long) of Ebola virus in West Africa.

% By quantifying errors in phylogenetic reconstruction we hope to persuade researchers looking to sequence pathogens during outbreaks of the utility of complete genome sequencing.
% We further generalise our results by relating four parameters - probability, evolutionary rate, alignment length and waiting times - that determine how much information can be encoded by evolving pathogen lineages via mutations, which can help guide future sequencing studies.
% \tbc{The use of `probability', etc... here is a bit opaque.'}
\end{abstract}

@@

\section*{Introduction}
The combination of decreasing cost of sequencing and the unparalleled insight it offers have led to the adoption of pathogen genetic sequencing as one of the most effective tools in a modern epidemiologist's toolkit.
When coupled with sophisticated models of evolution, pathogen sequences can be used to look into epidemiological features such as cryptic transmission \citep{faria_establishment_2017}, migration \citep{lemey_bayesian_2009,lemey_unifying_2014}, and origins \citep{smith_origins_2009} of infectious diseases, amongst others.
Pathogen sequences also contain information about past temporal dynamics before sequence data have been collected \citep{raghwani_origin_2012} due to the pattern of shared and unique mutations inherited from preceding generations.
Molecular phylogenetic approaches rely on decoding these patterns of shared mutations into a nested graph known as the phylogenetic tree.
Pathogens often have short generation times and some, like RNA viruses, also possess polymerases with low replication fidelity such that mutations are generated at a rapid pace \citep{drummond_measurably_2003,biek_measurably_2015} leading to fast differentiation of pathogen lineages at the genetic level as they spread.
With appropriate sampling and information (``metadata'') about sequences, historic population dynamics can be inferred and quantified from pathogen phylogenies.
Changes in pathogen population sizes over time \citep{pybus_integrated_2000}, inference of unobserved ancestral states \citep{lemey_bayesian_2009,dudas_mers-cov_2018}, correlates of processes \citep{faria_simultaneously_2013,lemey_unifying_2014,dudas_virus_2017}, and overall phylodynamic \citep{grenfell_unifying_2004} patterns can be inferred from molecular phylogenies and used to understand patterns of pathogen transmission at a number of scales.

Before widespread adoption of high-throughput sequencing limitations and costs led to amplification and sequencing of short fragments of pathogen genomes \citep{jin_genetic_1999,jin_proposal_2005}.
These subgenomic fragments were often chosen for their diversity, such as viral surface glycoproteins that experience selective pressures from vertebrate immune systems, or their utility, such as routine sequencing of HIV \textit{pol} gene to test for drug resistance \citep{kaye_phylogenetic_2008,rhee_human_2003}.
Whilst subgenomic fragments of pathogens are very accurate and specific as diagnostic markers and informative about long-term evolution, their length (dictated by the compromise between information content and ease of sequencing) limits their utility in detailed molecular epidemiology investigations, for example during outbreaks \citep{wohl_co-circulating_2018}, as only mutations occurring within the small region of the genome that is sequenced are available for phylogenetic inference.

Molecular clocks have been particularly useful in molecular epidemiology, where the accumulation of mutations between sequences is used as a noisy approximation for elapsed time, given either times of events in the phylogeny (sequence dates or dates of common ancestors) or a previously determined molecular clock rate.
% Mutations, prior to the action of selection, occur randomly across a genome via misincorporation of nucleotides during replication by the polymerase.
% Many of these mutations are filtered out by selection if they adversely affect protein function, with leftover mutations being beneficial, neutral, or mildly deleterious.
Generally, neutral pathogen variation at the nucleotide level ebbs and flows under the forces of population genetics, unlike beneficial or deleterious variation which tends to either fix or be purged rapidly, respectively.
Due to their random and discrete nature, mutations are modelled as a Poisson process \citep{yang_computational_2006} where the waiting time $t$ for observing a mutation at a single site is exponentially distributed with evolutionary rate parameter $R$.
The probability of observing 0 mutations at a single site after time $t$ is $e^{-Rt}$ and the probability of at least one mutation is therefore $1-e^{-Rt}$.
Higher evolutionary rates \textit{R} or waiting times \textit{t} result in higher probabilities of observing at least one mutation at the site in question.
Since sites are assumed to evolve independently, the probability of observing at least one mutation across \textit{L} sites is
\begin{equation}
  P = 1-e^{-RLt},
\end{equation}
where \textit{RL} is expressed in substitutions per year (rate in substitutions per site per year multiplied by number of sites).
Under a given evolutionary rate \textit{R}, and sequence length \textit{L}, we can rearrange the equation to quantify the mean waiting time for at least one mutation:
\begin{equation}
  t = \frac{1}{RL}
  \label{horizon}
\end{equation}
When the evolutionary rate \textit{R} or sequence length \textit{L} are low mean waiting times are lengthened and \textit{vice versa}.
Since both maximum plausible evolutionary rates and genome length are largely dictated by deleterious mutation load \citep{gago_extremely_2009}, neither quantity will vary substantially for a given pathogen, though individually $R$ and $L$ can vary substantially, where for example viruses have high $R$ and low $L$ on average, and bacteria have higher $L$ but lower $R$.
% Because of these natural constraints there is thus a lower limit to how many mutations can be expected to occur in a given length of sequence evolving at a particular evolutionary rate.
% We thus refer to the mean waiting time for a mutation as the temporal horizon, which reflects a theoretical temporal resolution over which mutations arise and are available for phylogenetic inference in a study system.

In this study we quantify how much information relevant to phylodynamic analysis is lost when shorter genomic regions are used instead of full genomes.
By focusing our attention on a subset (600 sequences) of a well-characterised genomic sequence data (comprised of >1600 viral genomes) set derived from the West African Ebola virus epidemic of 2013-2015 \citep{dudas_virus_2017} we estimate loss in precision and accuracy of molecular clock models and phylogeographic inference methods when only the glycoprotein gene (GP), a region representing just 10\% of the viral genome, is analysed despite GP evolving at rates faster than the genomic average.
Our methods rely on masking tip dates and locations for 60 (10\%) of the sequences in a classic training-testing split, where we re-infer these parameters as latent variables in MCMC.
We show that this reduction in data not only leads to severe convergence issues in Markov chain Monte Carlo (MCMC) analyses by removing the constraints additional data impose on plausible parameter space but can also result in unreliable tip date and location inference.
Despite achieving much better temporal resolution when using complete viral genomes we still find residual error caused by inherent randomness of mutations which is close to theoretical expectations (Eq \ref{horizon}).
We refer to this as the temporal horizon, \textit{i.e.} a temporal resolution limit where population processes occurring at a rate faster than the rate at which mutations enter and are observed in a population will not be captured with high fidelity, even with genome sequences.
% As such, pathogen genomes have an inherent temporal resolution limit, where processes occurring at a rate lower than the rate of evolution of a given pathogen will not be recovered reliably using currently standard phylogenetic methods.
% We refer to this as the temporal horizon and express it as the mean waiting time until at least one mutation occurs and differentiates a pair of evolving sequences.
 % which is already substantial for fine scale epidemiology and that sequencing subgenomic fragments of a pathogen results in loss of information that can make sequence data unreliable beyond diagnostics and typing.
% In order to highlight how much information is lost and to relate the concept of temporal horizon with real data we use a subset of a well-characterised dataset of Ebola virus genomes from West Africa collected in 2014-2015 \citep{dudas_virus_2017}.
% By analysing both complete viral genomes and the same dataset reduced to just the surface glycoprotein GP (corresponding with just 10\% of all genomic sites), we show substantial loss of predictive power, incorrect or unreliable reconstructions of past events, and migration model, despite increased evolutionary rate in GP compared to complete genome.
% We hope that analyses presented here may serve as a guide to researchers in the field of genetic epidemiology of how to think about the statistical power of sequence data itself, as well as encouragement to model developers to address the widening rift between data volumes and methods used to analyse them.

\begin{figure}[h]
 \centering
	\includegraphics[width=0.75\textwidth]{figures/fig1_trees.png}
	\caption{\textbf{Phylogenies of West African Ebola virus genomes (top) or GP sequences (bottom).}
	Temporal phylogenies recovered using BEAST are shown on the left, maximum likelihood phylogenies recovered with RAxML are on the right.
  Tips are coloured based on country (Sierra Leone in blue, Liberia in red, Guinea in green) and location (lighter colours indicate administrative divisions lying towards west of the country).
  Tips outlined in white indicate the 60 chosen for date and location masking, ticks to the right of phylogenies indicate the y positions of masked tips, coloured by their true location.
  In temporal phylogenies branches are also coloured based on GLM-inferred ancestral locations.
  Nodes in temporal phylogenies with <0.10 posterior probability are indicated with grey X marks.
  Maximum likelihood phylogenies on the right are rooted via temporal regression in TreeTime.
  Note that x axes between maximum likelihood phylogenies are not consistent, as GP evolves faster than the genomic average and thus has more substitutions per site.
	}
	\label{trees}
\end{figure}

@@

\subsection*{Loss of phylogenetic signal}
% In the past amplifying smaller fragments of pathogen genomes was often a pragmatic compromise between the cost of sequencing and information content, leading to sequencing of short but rapidly evolving pathogen genes or genomic regions.
% What few sequences were available were often collected months apart or across large geographic areas, ensuring that many sequences were genetically distinct either via mutations occurring in the intervening time or through prior population differentiation.
% However, in outbreak scenarios shorter genome fragments often do not accumulate mutations at a pace rapid enough to be of much use over short timescales.

Figure \ref{trees} shows the reconstructed phylogenies in substitution space (right) and time space (left) for 600 complete Ebola virus genomes (top) or just GP sequences (bottom).
Although higher levels of divergence are observed in the GP dataset (note that x axes between genome and GP phylogenies are not consistent), as seen from tree height, the differences in the number of non-polytomic nodes between genomic and GP data are clear, indicating substantially better resolution in disentangling the exact relationships between lineages in the former.
Figure \ref{embedding} shows where in the better resolved maximum likelihood phylogeny of genome sequences mutations that occurred in just the region spanning the GP gene.
Internal branches of a phylogeny correspond to hypotheses of common ancestry and in the case of GP only 42 internal nodes are identified in the maximum likelihood phylogeny compared to 210 internal nodes for complete genomes.
The few aspects of the West African epidemic that can be inferred from both GP and genome phylogenies, only the virus' origins in Guinea are clear, with details of its onwards spread largely lost in the GP phylogeny.
Genomic data, on the other hand, despite a reduction from over 1600 sequences described in the original study \citep{dudas_virus_2017} down to just 600, still contain information about the role of Sierra Leone's epidemic in maintaining transmission across the region through both endemic proliferation of lineages and their spread to neighbouring countries.

Unlike maximum likelihood phylogenies where branch lengths are directly proportional to the expected number of substitutions, branch lengths in temporal phylogenies are usually smoothed out by the fact that a range of dates are compatible with a given number of mutations on a branch.
Thus even large polytomies can be resolved into a branching structure derived from the tree prior, albeit without much support for any given configuration.
This can be seen in Figure \ref{trees}, where temporal phylogenies (left) appear to have similar degrees of resolution, yet the GP dataset (bottom) contains more nodes with less than 0.10 posterior support (marked by grey crosses).
Similarly, there is a noticeable degree of branch clustering by country in the GP temporal phylogeny, possibly caused by proximity of locations within country, which in the absence of genetic information cannot be resolved to the same degree as with genomic data.

In contrast to the maximum likelihood phylogeny of GP on the right (Figure \ref{trees}) its corresponding temporal phylogeny on the left exhibits a reconstruction of the West African epidemic largely consistent with what has been established previously \citep{dudas_virus_2017}.
This is likely to be caused by the combined effects of two sources of information.
First, additional information is added by specifying the collection dates for sequences, which might exclude certain topologies from being considered during MCMC on account of the relatively small effective population size of Ebola virus in West Africa.
Second, the generalised linear model approach to inferring migration is information-rich as it provides over 3000 possible parameter values (pairwise migration rates between locations) per predictor matrix, and thus if a few branches are strongly selecting for a ``correct'' predictor matrix to be included in the migration model that predictor matrix can then be used to determine the likely locations of branches for which less information is available.
However, a simpler continuous time Markov chain model where each individual pairwise migration rate is inferred individually in a maximum likelihood framework exhibits broadly similar patterns too (Figure \ref{TTtrees}).

\begin{figure}[h]
 \centering
	\includegraphics[width=0.6\textwidth]{figures/fig2_dates.png}
	\caption{\textbf{Masked tip date inference from genomes (left) and GP sequences (right).}
  Inferred collection dates in the masked set based on genomes (red, left) and GP sequences (blue, right).
  Each vertical line corresponds to the 95\% highest posterior density for the inferred tip date (y-axis), coloured red (genome) or blue (GP) if it falls within the true collection date (x-axis) and black otherwise.
  Dashed diagonal line indicates the 1-to-1 line.
  A histogram of residual errors (accuracy) between mean posterior date estimate and true date for each masked tip is shown in the second row.
  Third row shows the histogram of confidence interval widths for date estimates (precision).
	}
	\label{dates}
\end{figure}

@@

Inferring masked tip dates from 10\% of the sequences (Figure \ref{dates}) is an intuitive way to show both the inherent noisiness of molecular clock estimates, as reflected in the width of 95\% highest posterior density intervals for inferred dates, and the differences in temporal resolution between GP and genome alignments.
True collection dates for genomes are mostly (56 out of 60, corresponding to a coverage probability of 0.93) within the 95\% highest posterior density of estimated dates, and the mean absolute error is $\approx$22 days across all masked tips.
In contrast, the 95\% HPDs for inferred dates in the GP dataset capture more of the true dates (58 out of 60, coverage probability $\approx$0.96) at the cost of markedly reduced precision, with mean absolute error going up to $\approx$106 days or $\sim$3.5 months.
Despite having lower coverage probability, more precise date estimates are derived from complete genomes with an average 95\% HPD width of $\approx$102 days, compared to $\approx$458 days for GP.
Another way of thinking about where the loss of information occurs is to consider root-to-tip against tip date regressions shown in Figure \ref{rtt}, where waiting times for mutations are too long to estimate the slope of the regression reliably, as every new mutation is seen across sequences collected over a longer interval of time.
Observed errors (Figure \ref{dates}, but also Figure \ref{TTdates} for maximum likelihood equivalent) for both datasets are very close to theoretical expectations calculated using Equation \ref{horizon}: 22 (observed) versus 20 (expected) days for Ebola virus genomes, and 106 (observed) versus 113 (expected) days for GP.
Also note that for many tips in the GP data set independent Markov chains converged onto different dates for masked tips (\textit{i.e.} local maxima) resulting in multi-peaked posterior samples after combining independent analyses.

\begin{figure}[h]
 \centering
	\includegraphics[width=0.75\textwidth]{figures/fig3_locations.png}
	\caption{\textbf{Masked tip location inference from genomes (left) and GP sequences (right).}
  Horizontal bars indicate the posterior distribution of masked tip locations, coloured by country (Sierra Leone in blue, Liberia in red, Guinea in green) and location (lighter colours indicate administrative divisions lying towards west of the country).
  The correct location of each tip is outlined in white, with the smaller plot to the right showing only the posterior probability of the correct location.
  Bars marked with an asterisk indicate cases where the correct location is within the 95\% credible set.
  Posterior distribution of probability-weighted distances between population centroids of inferred and correct locations is shown at the bottom with mean indicated by a tick mark.
	}
	\label{locations}
\end{figure}

\begin{figure}[h]
 \centering
  \subfloat{
  \includegraphics[width=0.45\textwidth]{figures/figXA_histories.png}
  }
  \subfloat{
  \includegraphics[width=0.45\textwidth]{figures/figXB_histories.png}
  }
  \\
  \subfloat{
  \includegraphics[width=0.45\textwidth]{figures/figXC_histories.png}
  }
  \subfloat{
  \includegraphics[width=0.45\textwidth]{figures/figXD_histories.png}
  }
  \caption{\textbf{Posterior traces of ancestral locations and posterior migrations for four Ebola virus lineages from genomes (top) and GP sequences (bottom).}
  The inferred ancestral branch location is logged at time points along the path from selected tips to the root of the tree, across the posterior distribution of trees.
  The smoothed trajectories are an indication of where and when a lineage that gave rise to a particular tip is inferred to have existed.
  Maps on the right show migration events that are inferred to have taken place, coloured by their posterior probability, with migrations with <0.05 posterior support are shown as dotted white lines.
  All lineages trace back to Gu\'{e}ck\'{e}dou prefecture of Guinea (white outline in the map), where the original zoonotic transmission event occurred near the Guinean border with Sierra Leone and Liberia.
  Some lineages are also descended from an early spillover event into Sierra Leone.
  }
	\label{trace}
\end{figure}

\subsubsection*{Migration model is strongly informed by tip dates and locations}

Differences between genomic and GP datasets are clear and dramatic when looking at both phylogenies (Figure \ref{trees}) and masked date inference (Figure \ref{dates}), but less pronounced when trying to infer the location of a masked tip (Figure \ref{locations}).
Although locations are correctly inferred more often and with greater support in genomic sequences compared to just the GP gene, there are numerous tips whose locations are not correctly inferred even from genome sequences (Figure \ref{locations} and Figure \ref{TTlocations} for maximum likelihood equivalent).
This might reflect the nature of these parameters of interest, since phylogenies and date inference ultimately draw information from mutation accumulation via relatively straightforward models of sequence evolution with limited parameter space.
In contrast, migration processes are far more complicated and nuanced without a \textit{de facto} standard for modelling, though continuous time Markov chain (CTMC) approaches are widely used \citep{lemey_bayesian_2009} with most advanced methods relying on generalised linear models without excessive over-parameterisation \citep{faria_simultaneously_2013,lemey_unifying_2014,dudas_virus_2017}.
Despite the lack of strong contrast in power to infer masked locations between genomes and GP sequences, cross entropies indicate better performance with complete genomes (6054.631 nats) than with GP (9905.726 nats).

Similarly, locations are inferred correctly more often with complete genomes than with GP sequences, where the maximum probability location (\textit{i.e.} the model's best guess) matches the truth, where using complete genomes results in 0.540 probability of guessing correctly compared to 0.286 probability for GP (for a calibration of both models see Figure \ref{calibration}).
The model makes these guesses with more certainty too, where the mean probability of the true location is 0.482 with genomes and 0.219 with GP, and mean probability of best guess (\textit{i.e.} maximum probability) is 0.680 and 0.396, respectively.
% \tbc{Need to more carefully define what you mean by `inferred correctly'. Cross-entropy does this well, otherwise need a cut-off for assignment. Over 50\%?}
We also calculated the great circle distances between the population centroids of true and each predicted location weighed by probability, which should ideally be 0.0 (0 km distance multiplied by probability of 1.0).
The mean of these distances across masked tips are 75.886 kilometres for genomes compared to 164.309 km for GP sequences.
